---
title: "Canberra SA2 — School Aged Estimate Residential Population (5-14) Changes"
execute:
  echo: false
  message: false
  warning: false
---


```{r}
#| echo: false
#| message: false
#| warning: false
# Load libraries and data
library(sf)
library(leaflet)
library(dplyr)
library(crosstalk)
library(htmltools)
library(plotly)

erp <- st_read(paste0(here::here(), "/data/erp_sa2.gpkg"),
  quiet = T)

# Aggregate 5-9 and 10-14 age groups to create 5-14 school age population
erp <- erp |>
    filter(age %in% c("5-9", "10-14")) |>
    group_by(sa2_code, sa2_name, year, geom) |>
    summarise(erp = sum(erp, na.rm = TRUE), .groups = "drop") |>
    st_as_sf()

# Rename erp column to value for consistency with mapping code
erp <- erp |>
    mutate(value = erp, metric = "Population 5-14")
```

The below graphs demonstrate the estimated change in school aged population from 2001 to 2024. The few key things worth noting are that in 2001 there were a lot of young familes in south Canberra. In 2004, there are still a large number however, there has been a considerable decline. Instead, many more families have moved to the newer Canberra suburbs such as greater Gungahlin and Coombs/Wright area. Finally, overall there has been an increase of school aged children since 2001.

## Population Map

::: {.columns}

::: {.column width="50%"}

### 2001
```{r}
# Generated by Claude 3.5 Sonnet
# Simple approach - create individual leaflet maps for key years
#| echo: false
create_leaflet_map <- function(year_selected) {
  data_yr <- erp |> filter(year == year_selected)
  
  # Calculate color palette for this year
  pal <- colorNumeric("viridis", data_yr$value, na.color = "#cccccc")
  
  leaflet(data_yr) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    addPolygons(
      fillColor = ~pal(value),
      fillOpacity = 0.7,
      color = "#444444",
      weight = 1,
      label = ~paste0(sa2_name, ": ", value, " (", year, ")"),
      labelOptions = labelOptions(sticky = TRUE)
    ) |>
    addLegend(
      pal = pal, 
      values = ~value,
      title = paste("5-14 ERP <br>", year_selected, "")
    ) |>
    setView(lng = 149.12, lat = -35.30, zoom = 10)
}

create_leaflet_map(2001)

```
:::

::: {.column width="50%"}
### 2024
```{r}
#| echo: false
create_leaflet_map(2024)
```
:::

:::
# Change Analysis

### Absolute Change (2001-2024)
```{r}
#| echo: false
create_change_map <- function() {
  # Get 2001 and 2024 data
  data_2001 <- erp |> filter(year == 2001) |> select(sa2_code, value_2001 = value)
  data_2024 <- erp |> filter(year == 2024) |> select(sa2_code, value_2024 = value, sa2_name, geom)
  
  # Calculate absolute change
  change_data <- data_2024 |>
    left_join(st_drop_geometry(data_2001), by = "sa2_code") |>
    mutate(
      change = value_2024 - value_2001,
      change_label = ifelse(change >= 0, paste0("+", change), as.character(change)),
      fill_opacity = ifelse(value_2001 == 0 & value_2024 == 0, 0.3, 1.0)
    ) |>
    filter(!is.na(change))
  
  # Create diverging palette
  max_abs_change <- max(abs(change_data$change), na.rm = TRUE)
  pal <- colorNumeric(
    palette = c("#B2182B", "#D6604D", "#F4A582", "#FDDBC7", "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC"),
    domain = c(-max_abs_change, max_abs_change),
    na.color = "#cccccc"
  )
  
  leaflet(change_data) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    addPolygons(
      fillColor = ~pal(change),
      fillOpacity = ~fill_opacity,
      color = "#444444",
      weight = 1,
      label = ~paste0(
        sa2_name, "\n",
        "2001: ", value_2001, "\n", 
        "2024: ", value_2024, "\n",
        "Change: ", change_label
      ),
      labelOptions = labelOptions(sticky = TRUE)
    ) |>
    addLegend(
      position = "bottomright",
      pal = pal,
      values = ~change,
      title = "5-14 ERP Change\n2001-2024"
    ) |>
    setView(lng = 149.12, lat = -35.30, zoom = 10)
}

create_change_map()
```

### Population Trend (2001-2024)

```{r}
#| echo: false
library(ggplot2)
library(plotly)

create_population_trend <- function() {
  # Calculate total ACT population by year for 5-14 age group
  act_totals <- erp |>
    st_drop_geometry() |>
    group_by(year) |>
    summarise(total_population = sum(value, na.rm = TRUE), .groups = "drop")
  
  # Create ggplot
  p <- ggplot(act_totals, aes(x = year, y = total_population)) +
    geom_line(color = "#2166AC", linewidth = 1.2) +
    geom_point(color = "#2166AC", size = 3) +
    scale_x_continuous(breaks = seq(2001, 2024, 3)) +
    scale_y_continuous(labels = scales::comma_format()) +
    labs(
      title = "ACT School-Age Population Trend",
      subtitle = "Total population aged 5-14 years",
      x = "Year",
      y = "Population (5-14 years)"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 12),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )
  
  # Convert to plotly with custom hover info
  ggplotly(p, tooltip = "none") |>
    add_trace(
      data = act_totals,
      x = ~year, 
      y = ~total_population,
      type = "scatter",
      mode = "lines+markers",
      line = list(color = "#2166AC", width = 3),
      marker = list(color = "#2166AC", size = 8),
      hovertemplate = paste(
        "<b>Year:</b> %{x}<br>",
        "<b>Population:</b> %{y:,}<br>",
        "<extra></extra>"
      ),
      showlegend = FALSE
    ) |>
    layout(
      title = list(
        text = "ACT School-Age Population Trend<br><sub>Total population aged 5-14 years</sub>",
        font = list(size = 14)
      ),
      xaxis = list(title = "Year"),
      yaxis = list(title = "Population (5-14 years)", tickformat = ","),
      hovermode = "x unified"
    )
}

create_population_trend()
```
